The Evolution of Phenotypically Invariant Genetic Networks

### 2021-09-02 final_system_drift.tex
edits for final version

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses2}  % name of the review responses file

@@

{\centering
{\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
%Joshua S. Schiffman$^{\dagger}$ \footnote{Current affiliations: New York Genome Center, New York, New York 10013, U.S.A. \& Weill Cornell Medicine, New York, New York 10065, U.S.A.} \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
%$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
%$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
%{\small \texttt{jschiffman@nygenome.org} \qquad 
%\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
\small \today \\
\vspace{0.25cm}
}

% Other title ideas:
%
% Rapid speciation despite conservation of phenotype
%
% How fast does network drift create incompatibilities?
%
% More than one way to grow a cat: speciation despite conservation of phenotype
%
% More than one way to grow a cat: regulatory network drift and speciation
%
% More than one way to grow a cat: regulatory network redundancy and speciation
%
% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
%
% More than one way to grow a cat: can network drift lead to speciation?
%
% Beyond the snowball: a quantitative model of incompatibility accumulation
%
% System drift and speciation: more than one way to grow a cat
%
% An explicit model of neutral regulatory network evolution, with applications
% to the rate of accumulation of hybrid incompatibility
%
% Evolutionary conservation of phenotype does not entail conservation of the underlying
% molecular mechanism leading to rapid speciation
%
% The evolution of phenotype-invariant gene networks rapidly leads to hybrid incompatibiliy
%
% Evolutionary network rewiring can rapidly lead to hybrid incompatibility despite
% phenotypic conservation
%
% Hybrid incompatibilities can evolve rapidly due to developmental system drift
%
% Evolutionary systems theory and speciation

\begin{abstract}
%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype does not change over evolutionary time, 
the underlying mechanism may change, 
as distinct molecular pathways can realize identical phenotypes.
Here we use linear system theory to explore the consequences of this idea,
describing how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
This characterization implies that selectively neutral directions in genotype space should be common
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of independently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
is expected to proceed at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this could lead to substantial interpopulation incompatibility,
and thus speciation, on a time scale of $N_e$ generations.
This model also naturally predicts Haldane's rule, 
%thus providing another possible explanation
thus providing a concrete explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes 
during the early stages of speciation.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}

@@

At all levels of biological organization,
the problems that biological systems have evolved to solve
often do not have single solutions --
systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits.
In many cases, these functionally equivalent structures
can be explored through small, local changes to the structure that leave the function unchanged.
For instance, there are ``neutral networks''
of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}
amino acid sequences that fold similarly \citep{babajide1997neutral},
or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
Further examples are found in the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
% \plr{'structural plasticity' means something specific in neuroscience}

@@

In this paper we use results on mathematical nonidentifiability from \llabel{r21}
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
%
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first}.

% It is not a new observation
The field of population genetics has also explored the consequences of the fact
%that phenotype does not prescribe a unique genotype
that there is often more than one way to do the same thing, \revpoint{1}{2}
\llabel{r22}
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. 
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. 
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
%
Our main aim is to provide a concrete framework 
that can make natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

We use a model of gene regulatory networks
that describes the temporal dynamics of
a collection of $n$ coregulating molecules
-- such as transcription factors -- as well as external or environmental inputs.
% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
% within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$,
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$
into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions on the input
(e.g., $|u(t)|$ is integrable) 
we can write the phenotype as 
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{(t-s) A} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
The $i^\text{th}$ row of $A$ is therefore determined by genetic features such as
the strength of $j$-binding sites in the promoter of gene $i$,
factors affecting chromatin accessibility near gene $i$,
or basal transcription machinery activity.
The form of $B$ determines how the environment influences transcription factor expression levels,
and $C$ might determine the rate of production of downstream enzymes.
%% {Linearity} also ``zero-state equivalence''
% Here we have assumed that the system is linear,
% and begins from the ``zero'' state ($\kappa(0)=0$).
% Of course, neither of these are necessarily true for real systems,
% but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
% Furthermore, the ease of analyzing linear systems makes this an attractive place to start.

@@

\begin{example}[An oscillator]\label{ex:oscillator} \llabel{r11b}
  For illustration, we consider an extremely simplified model of oscillating gene transcription,
as for instance is found in cell cycle control or the circadian rhythm.
% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
There are two genes, 
whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
Only the dynamics of gene-1 are consequential to the oscillator 
(perhaps the amount of gene-1 activates another downstream gene network). 
Lastly, both genes are equally upregulated by an exogenous signal.
The dynamics of the system are described by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
        \phi(t) &= \kappa_{1}(t)   .
    \end{align*}
In matrix form the system regulatory coefficients are given as,
$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}
If the input is an impulse at time zero (a delta function), 
then the phenotype is equal to the impulse response: 
\begin{align*}
\phi(t) = h(t) = \sin t + \cos t  .
\end{align*}
The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
\end{example}

\begin{figure}[H]
   % \begin{center} 
  \centering
     \begin{tabular}{cc}
        \begin{tikzpicture}
        \begin{scope}[every node/.style={circle,thick,draw}]
            \node (A) at (0,0) {$\kappa_{1}$};
            \node (B) at (4,0) {$\kappa_{2}$};
            \node[shape=rectangle] (U) at (2,2) {input ($u$)};
            \node[shape=rectangle] (y) at (2,-2) {output ($\phi$)};
            %\\  \includegraphics[width=0.15\textwidth, height=0.04\paperheight]{osc_impulse} \end{tabular}};
        \end{scope}

        \begin{scope}[>={Stealth[black]},
                      every node/.style={fill=white,circle},
                      every edge/.style={draw=black, thick}]
            \path [->] (A) edge[bend left] node {\tiny $-1$} (B);
            \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
            \path[->] (U) edge node {\tiny $1$} (A);
            \path[->] (U) edge node {\tiny $1$} (B);
            \path[->] (A) edge[bend right] node {\tiny $1$} (y);
        \end{scope}
        \begin{scope}[>={Stealth[black]},
                      every edge/.style={draw=black, thick}]
            %\path [->] (A) edge[loop left] node {\tiny $\lambda_{1}$} (A);
            %\path [->] (B) edge[loop left] node {\tiny $\lambda_{2}$} (B);
        \end{scope}

     \end{tikzpicture} &
     \includegraphics{examples/fig1b_small}
\end{tabular}
%  \end{center}
  \caption{
    (Left) 
    Diagram of the gene network in Example \ref{ex:oscillator}
  , and (right) plot of the phenotype $\phi(t)$ against time $t$
  .} \label{fig:oscillator}
\end{figure}

@@

Two systems are equivalent if they produce the same phenotype given the same input,
i.e., have the same input--output relationship.
We say that
the systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
\textbf{phenotypically equivalent} 
if their impulse response functions are the same:
$h(t) = \bar h(t)$ for all $t \ge 0$.
This implies that for any acceptable input $u(t)$,
if $(\kappa_u(t),\phi_u(t))$ and $(\bar \kappa_u(t),\bar \phi_u(t))$ are the solutions to equation \eqref{eqn:system}
of these two systems, respectively, then
\begin{align*}
      \phi_u(t) = \bar \phi_u(t) \qquad \text{for all} \; t \ge 0.
\end{align*}
% (recall that $\kappa(0) = \bar \kappa(0) = 0$). 
In other words, phenotypically equivalent systems respond identically for \emph{any} input.

@@

This implies that most systems have at least $n^2$ degrees of freedom,
where recall $n$ is the number of components of the kryptotype vector.
This is because for an arbitrary $n \times n$ matrix $Z$,
taking $V$ to be the identity matrix plus a small perturbation in the direction of $Z$
above implies that
moving $A$ in the direction of $ZA - AZ$
% (e.g., $\bar A = A + \epsilon(ZA-AZ)$)
while also moving $B$ in the direction of $ZB$ 
and $C$ in the direction of $-CZ$
will leave the phenotype unchanged to second order in the size of the perturbation.
%% I believe the following is necessary and sufficient:
If the columns of $B$ and the rows of $C$ are not all eigenvectors of $A$,
then any such $Z$ will result in a different system.

It turns out that in general, there are more degrees of freedom,
except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
to achieve the desired dynamics.
Results in system theory show that any time-invariant \revpoint{2}{5}
system can be realized in a particular minimal dimension
(the dimension of the kryptotype, $n_\text{min}$),
and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
Since gene networks can grow or shrink following gene duplications and deletions, 
these additional degrees of freedom can apply, in principle, to any system.

Even if the system is not minimal, results from systems theory % -- the Kalman decomposition --
explicitly describe the set of all phenotypically equivalent systems.
% Could define script S here for 'system' if we want: $\Sys$
We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
to the system defined by $(A_0, B_0, C_0)$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \allS(A_0, B_0, C_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  .
  \end{aligned}
\end{equation}
These systems need not have the same kryptotypic dimension $n$,
but must have the same input and output dimensions ($\ell$ and $m$, respectively).

The Kalman decomposition, which we now describe informally, elegantly characterizes this set
\citep{kalman1963mathematical,kalman1969topics,anderson1966equivalence}.
To motivate this, first note that the input $u(t)$ only directly pushes the system
in certain directions (those lying in the span of the columns of $B$).
As a result, different combinations of input can 
move the system in any direction that lies in what is known as the \emph{reachable subspace}.
Analogously, we can only observe motion of the system in certain directions
(those lying in the span of the rows of $C$),
and so can only infer motion in what is known as the \emph{observable subspace}.
The Kalman decomposition then classifies each direction in kryptotype space\llabel{r23}
as either reachable or unreachable, and as either observable or unobservable.
Only the components that are both reachable and observable determine the system's phenotype --
that is, components that both respond to an input and produce an observable output. 
% it is possible to add components that respond to an input, 
% but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs.

@@

\includereviews

>>>>

% Responses to reviews
%\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
%\newif\ifreviewresponses
%\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
%\newcommand{\responsefile}{review-responses2}  % name of the review responses file

@@

{\centering
{\Huge \bf System drift and speciation} \\ \vspace{0.75cm}
Joshua S. Schiffman$^{\dagger}$ \footnote{Current affiliations: New York Genome Center, New York, New York 10013, U.S.A. \& Weill Cornell Medicine, New York, New York 10065, U.S.A.} \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\ \vspace{0.5cm}
$^{\dagger}${\footnotesize {Molecular and Computational Biology, University of Southern California, Los Angeles, California 90089, U.S.A. \\
$^{\ddagger}$Departments of Mathematics and Biology \& The Institute for Ecology and Evolution, University of Oregon, Eugene, Oregon 97403, U.S.A.}} \\ \vspace{0.5cm}
{\small \texttt{jschiffman@nygenome.org} \qquad 
\texttt{plr@uoregon.edu}} \\ \vspace{0.5cm}
\small \today \\
\vspace{0.25cm}
}

\begin{abstract}
Even if a species' phenotype does not change over evolutionary time, 
the underlying mechanism may change, 
as distinct molecular pathways can realize identical phenotypes.
Here we use linear system theory to explore the consequences of this idea,
describing how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small changes to these molecular pathways
causes a population to explore the set of mechanisms with identical phenotypes.
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
This characterization implies that selectively neutral directions in genotype space should be common
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of independently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
is expected to proceed at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this could lead to substantial interpopulation incompatibility,
and thus speciation, on a time scale of $N_e$ generations.
This model also naturally predicts Haldane's rule, 
thus providing a concrete explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes 
during the early stages of speciation.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

@@

At all levels of biological organization,
the problems that biological systems have evolved to solve
often do not have single solutions --
systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself 
all the way up to the convergent evolution of adaptive traits.
In many cases, these functionally equivalent structures
can be explored through small, local changes to the structure that leave the function unchanged.
For instance, there are ``neutral networks''
of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}
amino acid sequences that fold similarly \citep{babajide1997neutral},
or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
Further examples are found in the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}.

@@

In this paper we use results on mathematical nonidentifiability from %\llabel{r21}
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
%
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first}.

The field of population genetics has also explored the consequences of the fact
that there is often more than one way to do the same thing, %\revpoint{1}{2}
%\llabel{r22}
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. 
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. 
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
%
Our main aim is to provide a concrete framework 
that can make natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

We use a model of gene regulatory networks
that describes the temporal dynamics of
a collection of $n$ coregulating molecules
-- such as transcription factors -- as well as external or environmental inputs.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$,
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$
into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions on the input
(e.g., $|u(t)|$ is integrable) 
we can write the phenotype as 
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{(t-s) A} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
The $i^\text{th}$ row of $A$ is therefore determined by genetic features such as
the strength of $j$-binding sites in the promoter of gene $i$,
factors affecting chromatin accessibility near gene $i$,
or basal transcription machinery activity.
The form of $B$ determines how the environment influences transcription factor expression levels,
and $C$ might determine the rate of production of downstream enzymes.

@@

\begin{example}[An oscillator]\label{ex:oscillator} %\llabel{r11b}
  For illustration, we consider an extremely simplified model of oscillating gene transcription,
as for instance is found in cell cycle control or the circadian rhythm.
There are two genes, 
whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
Only the dynamics of gene-1 are consequential to the oscillator 
(perhaps the amount of gene-1 activates another downstream gene network). 
Lastly, both genes are equally upregulated by an exogenous signal.
The dynamics of the system are described by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
        \phi(t) &= \kappa_{1}(t)   .
    \end{align*}
In matrix form the system regulatory coefficients are given as,
$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
If the input is an impulse at time zero (a delta function), 
then the phenotype is equal to the impulse response: 
\begin{align*}
\phi(t) = h(t) = \sin t + \cos t  .
\end{align*}
The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
\end{example}

        \begin{scope}[>={Stealth[black]},
                      every node/.style={fill=white,circle},
                      every edge/.style={draw=black, thick}]
            \path [->] (A) edge[bend left] node {\tiny $-1$} (B);
            \path [->] (B) edge[bend left] node {\tiny $1$} (A); 
            \path[->] (U) edge node {\tiny $1$} (A);
            \path[->] (U) edge node {\tiny $1$} (B);
            \path[->] (A) edge[bend right] node {\tiny $1$} (y);
        \end{scope}
        \begin{scope}[>={Stealth[black]},
                      every edge/.style={draw=black, thick}]
        \end{scope}

     \end{tikzpicture} &
     \includegraphics{examples/fig1b_small}
\end{tabular}
  \caption{
    (Left) 
    Diagram of the gene network in Example \ref{ex:oscillator}
  , and (right) plot of the phenotype $\phi(t)$ against time $t$
  .} \label{fig:oscillator}
\end{figure}

@@

Two systems are equivalent if they produce the same phenotype given the same input,
i.e., have the same input--output relationship.
We say that
the systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
\textbf{phenotypically equivalent} 
if their impulse response functions are the same:
$h(t) = \bar h(t)$ for all $t \ge 0$.
This implies that for any acceptable input $u(t)$,
if $(\kappa_u(t),\phi_u(t))$ and $(\bar \kappa_u(t),\bar \phi_u(t))$ are the solutions to equation \eqref{eqn:system}
of these two systems, respectively, then
\begin{align*}
      \phi_u(t) = \bar \phi_u(t) \qquad \text{for all} \; t \ge 0.
\end{align*}
In other words, phenotypically equivalent systems respond identically for \emph{any} input.

@@

This implies that most systems have at least $n^2$ degrees of freedom,
where recall $n$ is the number of components of the kryptotype vector.
This is because for an arbitrary $n \times n$ matrix $Z$,
taking $V$ to be the identity matrix plus a small perturbation in the direction of $Z$
above implies that
moving $A$ in the direction of $ZA - AZ$
while also moving $B$ in the direction of $ZB$ 
and $C$ in the direction of $-CZ$
will leave the phenotype unchanged to second order in the size of the perturbation.
If the columns of $B$ and the rows of $C$ are not all eigenvectors of $A$,
then any such $Z$ will result in a different system.

It turns out that in general, there are more degrees of freedom,
except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
to achieve the desired dynamics.
Results in system theory show that any time-invariant %\revpoint{2}{5}
system can be realized in a particular minimal dimension
(the dimension of the kryptotype, $n_\text{min}$),
and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
Since gene networks can grow or shrink following gene duplications and deletions, 
these additional degrees of freedom can apply, in principle, to any system.

Even if the system is not minimal, results from systems theory % -- the Kalman decomposition --
explicitly describe the set of all phenotypically equivalent systems.
We refer to $\allS(A_0,B_0,C_0)$ as the set of all systems phenotypically equivalent
to the system defined by $(A_0, B_0, C_0)$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \allS(A_0, B_0, C_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  .
  \end{aligned}
\end{equation}
These systems need not have the same kryptotypic dimension $n$,
but must have the same input and output dimensions ($\ell$ and $m$, respectively).

The Kalman decomposition, which we now describe informally, elegantly characterizes this set
\citep{kalman1963mathematical,kalman1969topics,anderson1966equivalence}.
To motivate this, first note that the input $u(t)$ only directly pushes the system
in certain directions (those lying in the span of the columns of $B$).
As a result, different combinations of input can 
move the system in any direction that lies in what is known as the \emph{reachable subspace}.
Analogously, we can only observe motion of the system in certain directions
(those lying in the span of the rows of $C$),
and so can only infer motion in what is known as the \emph{observable subspace}.
The Kalman decomposition then classifies each direction in kryptotype space %\llabel{r23}
as either reachable or unreachable, and as either observable or unobservable.
Only the components that are both reachable and observable determine the system's phenotype --
that is, components that both respond to an input and produce an observable output.

@@

%\includereviews
